### 2017-03-20 Manuscript/Cover letter.Rmd
MAJOR EDITS
Changed to pdf format
Edits - submitted!

School of Biological Sciences\\*
University of East Anglia\\*
Norwich\\*
Norfolk NR4 7TJ\\* 
United Kingdom

@@

@@

Secondly, using a large number of repeated telomere measurements we show that within-individual increases in telomere length cannot be solely explained by measurement error. This is in accordance with recent theoretical work, and data from humans, suggesting that telomere elongation may be an ecologically important process. We hope that our data will spur further research into when, where any why telomere elongation occurs.

>>>>

Dr Lewis G. Spurgin\\*
BBSRC Research Fellow\\*
School of Biological Sciences\\*
University of East Anglia\\*
Norwich\\*
Norfolk, NR4 7TJ\\* 
United Kingdom

@@

\bigskip
\bigskip

@@

Secondly, using a large number of repeated telomere measurements we show that within-individual increases in telomere length occure, and that these cannot be solely explained by measurement error. This is in accordance with recent theoretical work, and data from humans, suggesting that telomere elongation may be an ecologically important process. We hope that our data will spur further research into when, where any why telomere elongation occurs.

### 2017-03-20 Manuscript/SW_early_life_telomeres_FULL.Rmd
Edits - submitted!

*Study species and sampling*
The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca* 320 adult birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Komdeur1992; @Richardson2003b; @Spurgin2014a]. This species' main breeding season runs from June-September (though a small proportion of pairs also breed between January–March), when the breeding females on many of the *ca* 110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991]. Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50). The 22 birth year cohorts used in this study cover 1993 to 2014 – the time period during which our data and sampling are most complete.

@@

*Molecular methods*
For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l. Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998]. Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific). The following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 absorbance ratio has to be between 1.8 and 2.0 for acceptable DNA purity, and iii) the 260/230 absorbance ratio must be higher than 1.8. DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded. All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, following Bebbington *et al.* [-@Bebbington2016]. Based on the distribution of observed cq values, we excluded outlier samples with extremely large cq values (cq values > 25 and 26 were excluded for the telomere and GAPDH reactions, respectively), which were assumed to be failed reactions.

@@

*Statistical analyses*
We performed all statistical analyses using R version 3.2.2 [@RDevelopmentCoreTeam2011]. RTL was square root transformed to improve linear model fits, and we assessed repeatability of RTL using the rptR package.

@@

*Telomere dynamics and age among cohorts*
We first tested how RTL was related to age among cohorts using a model selection approach. The top model contained a log-linear relationship between RTL and age (Table 1A). All other models fitted the data much less well ($\Delta$AICc > 15; Table 1). The log-linear relationship between RTL and age could be seen clearly in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring in the first year of life (Fig. 1A). There was substantial variation in RTL among cohorts, with no obvious trend over time (Fig. 1B). There was a negative relationship between RTL and log age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope the relationship varied substantially among cohorts (Fig. 1C). To test whether this variation was significant we fitted a model including the log age x cohort interaction term, and found that this was a marginally better fit than a model including only main effects ($\Delta$AICc = `r round2(AICc(RTL_LogAge) - AICc(RTL_LogAgeXcohort))`). In the one year in which RTL increased with age (2013), 17 of the 18 birds sampled were fledglings or subadults, suggesting that the observed pattern was an artefact of the sampling in this season (i.e. a lack of variation in age among sampled birds), rather than a real relationship.

@@

*Telomere dynamics and the environment*
In addition to age, RTL was associated with tarsus length, sex and insect abundance (Fig. 3A). RTL was negatively related to tarsus length and males had longer telomeres than females (Fig. 3B), while insect abundance was positively related to RTL (Fig. 3C). The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`). The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig. S1). One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1). In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL decreased with tarsus length in both sexes, but this decrease was stronger in females (Fig. 3B). No social or ecological environmental variables were significant predictors of $\Delta$RTL using the full model approach (Table S2). Using model selection, we found that the top model explaining $\Delta$RTL contained age and population density (Table S3). $\Delta$RTL was positively related to age, consistent with telomere shortening being highest in early life, and negatively related to population density; however, in both instances model averaged confidence intervals overlapped zero (Fig. S2).

@@

nams[nams == 'DeltaAge'] <- 'Delta age (linear)'
nams[nams == 'DeltaLogAge'] <- 'Delta age (log)'
nams[nams == 'Cohort:DeltaAge'] <- 'Delta age (continuous)*cohort'
nams[nams == 'Cohort:DeltaLogAge'] <- 'Delta age (log)*cohort'
nams[nams == 'Cohort:DeltaAge2'] <- 'Delta age (quadratic)*cohort'
nams[nams == 'DeltaAge2'] <- 'Delta age (quadratic)'

>>>>

*Study species and sampling*

The Seychelles warbler is a small (~15 g), insectivorous passerine bird with a mean life expectancy of 5.5 years at fledging [@Hammers2013]. The population of *ca* 320 adult birds on Cousin Island (04'20'S, 55'40'E) has been intensively studied since 1986 [@Komdeur1992; @Richardson2003b; @Spurgin2014a]. This species' main breeding season runs from June-September (though a small proportion of pairs also breed between January–March), when the breeding females on many of the *ca* 110 territories will attempt to breed, laying one or, rarely, two or three eggs [@Komdeur1991]. Breeding attempts are often unsuccessful, and as a result of this low reproductive output, and higher mortality in first-year birds [39% in first-year birds versus 16% in adults; @Brouwer2006], cohort sizes in the Seychelles warbler are typically small (< 50). The 22 birth year cohorts used in this study cover 1993 to 2014 – the time period during which our data and sampling are most complete.

@@

*Molecular methods*

For each sample, genomic DNA was extracted from a ~2 mm^2^ flake of preserved blood using the DNeasy Blood and Tissue Kit (Qiagen), following the manufacturer's protocol, with the modification of overnight lysis at 37^o^C and a final DNA elution volume of 80 $\mu$l. Sex was determined using the PCR-based method outlined by Griffiths *et al.* [-@Griffiths1998]. Prior to telomere analysis, DNA concentration and purity were quantified using a NanoDrop 8000 Spectrophotometer (ThermoScientific). The following thresholds were applied before samples were included for further analysis: i) DNA concentration must be at least 15 ng $\mu$l^-1^ (based on a mean of three measurements), ii) the 260/280 absorbance ratio has to be between 1.8 and 2.0 for acceptable DNA purity, and iii) the 260/230 absorbance ratio must be higher than 1.8. DNA integrity was further validated by visualization with ethidium bromide after electrophoresis on a 1.2% agarose gel, and all samples with evidence of DNA degradation were re-extracted or excluded. All DNA extractions that passed the above criteria were diluted to 3.3 ng $\mu$l^-1^ before telomere measurement. We measured relative telomere length (RTL) for all samples using a quantitative PCR (qPCR) assay of telomeres and a GAPDH control gene, following Bebbington *et al.* [-@Bebbington2016]. Based on the distribution of observed cq values, we excluded outlier samples with extremely large cq values (cq values > 25 and 26 were excluded for the telomere and GAPDH reactions, respectively), which were assumed to be failed reactions.

@@

*Statistical analyses*

We performed all statistical analyses using R version 3.2.2 [@RDevelopmentCoreTeam2011]. RTL was square root transformed to improve linear model fits, and we assessed repeatability of RTL using the rptR package.

@@

*Telomere dynamics and age among cohorts*

We first tested how RTL was related to age among cohorts using a model selection approach. The top model contained a log-linear relationship between RTL and age (Table 1A). All other models fitted the data much less well ($\Delta$AICc > 15; Table 1). The log-linear relationship between RTL and age could be seen clearly in the raw data; RTL decreased with age (estimate = `r round3(fixef(RTL_LogAge)['LogAge'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_LogAge)['LogAge',],lessthan = F),collapse = ', ')`), with the greatest decrease occurring in the first year of life (Fig. 1A). There was substantial variation in RTL among cohorts, with no obvious trend over time (Fig. 1B). There was a negative relationship between RTL and log age in 21 of the `r length(table(dd$Cohort))` cohorts, but the slope the relationship varied substantially among cohorts (Fig. 1C). To test whether this variation was significant we fitted a model including the log age x cohort interaction term, and found that this was a marginally better fit than a model including only main effects ($\Delta$AICc = `r round2(AICc(RTL_LogAge) - AICc(RTL_LogAgeXcohort))`). In the one year in which RTL increased with age (2013), 17 of the 18 birds sampled were fledglings or subadults, suggesting that the observed pattern was an artefact of the sampling in this season (i.e. a lack of variation in age among sampled birds), rather than a real relationship.

@@

*Telomere dynamics and the environment*

In addition to age, RTL was associated with tarsus length, sex and insect abundance (Fig. 3A). RTL was negatively related to tarsus length and males had longer telomeres than females (Fig. 3B), while insect abundance was positively related to RTL (Fig. 3C). The full model was weak in terms of explanatory power of fixed effects (marginal R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2m'])`), although including the random effect terms increased this substantially (conditional R^2^ = `r round2(r.squaredGLMM(RTL_Full)['R2c'])`). The model averaging approach yielded qualitatively identical results to the full LMM, with the same explanatory variables 'significant' in terms of being retained in top models, and having model-averaged confidence intervals not overlapping zero (Table S1; Fig. S1). One interesting finding from the model selection was that sex only appeared in top models where tarsus length was also present (Table S1). In accordance with this, when tarsus length was removed from the full model sex was no longer significant (estimate = `r round3(fixef(RTL_Full_NoTarsus)['SexMales'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_NoTarsus)['SexMales',],lessthan = F),collapse = ', ')`), and a sex x tarsus interaction was significant when included (estimate = `r round3(fixef(RTL_Full_SexTarsus)['SexMales:Tarsus'],lessthan = F)`, CIs = `r paste(round3(CIwald(RTL_Full_SexTarsus)['SexMales:Tarsus',],lessthan = F),collapse = ', ')`); RTL decreased with tarsus length in both sexes, but this decrease was stronger in females (Fig. 3B). No social or ecological environmental variables were significant predictors of $\Delta$RTL using the full model approach (Table S2). Using model selection, we found that the top model explaining $\Delta$RTL contained age and population density (Table S3). $\Delta$RTL was positively related to age, consistent with telomere shortening being highest in early life, and negatively related to population density; however, in both instances model averaged confidence intervals overlapped zero (Fig. S2).

@@

nams[nams == 'DeltaAge'] <- 'Delta age (linear)'
nams[nams == 'DeltaLogAge'] <- 'Delta age (log)'
nams[nams == 'Cohort:DeltaAge'] <- 'Delta age (continuous)*cohort'
nams[nams == 'Cohort:DeltaLogAge'] <- 'Delta age (log)*cohort'
nams[nams == 'Cohort:DeltaAge2'] <- 'Delta age (quadratic)*cohort'
nams[nams == 'DeltaAge2'] <- 'Delta age (quadratic)'
nams[nams == 'MeanAge'] <- 'Mean age'
